## NYC Flights in 2013

For these exercises, we will use the dataframe `flights` from the `nycflights13` package, which has one row per flight that departed from New York City in 2013 (336776 rows).

```{r}
library(tidyverse)

if (!("nycflights13" %in% installed.packages())) {
  install.packages("nycflights13")
}
library(nycflights13)
flights <- flights
```

These exercises come from (or were adapted from) [R for Data Science](https://r4ds.hadley.nz/data-transform.html#exercises).

Hint: Exercises 1 through 4 can be solved with `filter()`, `distinct()`, and `count()` from `dplyr`.

1. Was there a flight on every day in 2013?

```{r}
flights |>
  distinct(day, month) %>%
  count()
```


2. Find all flights that arrived more than two hours late but didn’t leave late.

```{r}
flights |>
  filter(dep_delay <= 0) |>
  filter(arr_delay >= 120)
```

3. Find all destinations that had at least one flight with a flight time of 12 hours or more.

```{r}
flights |>
  filter(air_time >= 600) |>
  distinct(dest)
```

4. Find all flights in the summer (June, July, August) that arrived on time.

```{r}
flights |>
  filter(month >= 6, month <= 8) |>
  filter(arr_delay <= 0)
```